---
title: "Load_GSE48024"
author: "Sylvia Yu Zhang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r}
# HOME PATH TO CHANGE
dir <- "/pine/scr/e/u/eunchong/BIOS784/BIOS784_FinalProject"
setwd(dir)
```


```{r load}
library(GEOquery)

#library(limma) # Not needed for loading data

#library(umap) # Not needed for loading data

library(knitr) # Needed to render RMD.
```

### Load data.
```{r}
gset <- getGEO("GSE48024", GSEMatrix =TRUE, AnnotGPL=TRUE)
```

```{r save_data}
saveRDS(gset, file=file.path(dir,"data","gset"))
load(file.path(dir,"data","gset"))
```


Here, "gset" contains two items, one data for male and another data for female. In total, there are 417 samples from 110 female participants, and 431 samples from 166 male participants. 

```{r}
gset_female <- gset[[1]]
gset_female
gset_male <- gset[[2]]
gset_male

# Information given regarding each sample
colnames(gset_female@phenoData)

table(gset_female$`gender:ch1`)
length(unique(gset_female$`subject:ch1`))

table(gset_male$`gender:ch1`)
length(unique(gset_male$`subject:ch1`))
```

Apparently, female and male samples are sequenced using difference version of platforms. Thus, we see different number of probes. 

```{r}
table(gset_female$platform_id) # Illumina HumanHT-12 V4.0 expression beadchip
nrow(gset_female)

table(gset_male$platform_id) # Illumina HumanHT-12 V3.0 expression beadchip
nrow(gset_male)
```

"AnnotGPL" option in "getGEO" function automatically adds annotation corresponding to each gene. However, 16033 probes do not have gene symbols annotated. 

```{r}
# Information given regarding each probe
colnames(gset_female@featureData)

kable(head(data.frame(ID = gset_female@featureData$"ID",
                      Gene_symbol = gset_female@featureData$"Gene symbol")))

kable(head(data.frame(ID = gset_female@featureData$"ID",
                      GO = gset_female@featureData$"GO:Function")))


sum(gset_female@featureData$"Gene symbol" == "")

```

### Check normalization, and expression range.

Just randomly draw 10 samples for each gender and boxplot each sample. 
```{r}
boxplot(exprs(gset_female)[,sample(1:417, 10, replace=F)], main = "Female")

boxplot(exprs(gset_male)[,sample(1:431, 10, replace=F)], main = "Male")
```

### Number of samples at each time for female and male

```{r}
kable(table(gset_female$`time:ch1`))

kable(table(gset_male$`time:ch1`))
```

```{r}
table(gset_female@phenoData@data$`time:ch1`,
      gset_female@phenoData@data$`treatment:ch1`)

table(gset_male@phenoData@data$`time:ch1`,
      gset_male@phenoData@data$`treatment:ch1`)
```

```{r}
class(gset_female)
```

```{r}
library(tidySummarizedExperiment)

se_female <- SummarizedExperiment(assay = list("exprs" = exprs(gset_female)),
                                  colData = as(gset_female@phenoData, "data.frame"),
                                  rowData = as(gset_female@featureData, "data.frame"))



se_male <- SummarizedExperiment(assay = list("exprs" = exprs(gset_male)),
                                  colData = as(gset_male@phenoData, "data.frame"),
                                  rowData = as(gset_male@featureData, "data.frame"))
```

The code below seeks to find genes with greatest amount of expression on Day0
```{r}
se_female_Day0 <- se_female %>% filter(`time:ch1` == "Day0")

hist(assay(se_female_Day0), xlab = "Amount of Expression", main = "Female")

dim(assay(se_female_Day0))
qx <- summary(as.numeric(assay(se_female_Day0)))
qx[3]

se_female_Day0 %>% 
  filter(exprs > qx[3])

a1 <- se_female_Day0 %>% 
  filter(exprs > qx[3])

a2 <- a1 %>%
  filter(`Gene ID` == "") %>%
  select(Platform_SEQUENCE)
# confirmed that the Platform_SEQUENCE with gene_id = '' is not associated with a gene based on current Illumina annotation info in https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL10558

a1 %>%
  count(`Gene ID`) %>%
  filter(!(`Gene ID`=="")) %>%
  dplyr::arrange(-n)

# not done
# then need to assess those top 5/10 genes' function, properties...
se_female_Day0 %>% filter(exprs > qx[3]) %>% select(`GO:Function`)
```


### Below shows five probes corresponding to gene "IL18BP", annotation done by "AnnotGPL" option.

I never tried annotation using "AnnotGPL" before, so I'm not sure how it is done exactly. 

```{r}
gset_female@featureData$"ID"[which(gset_female@featureData$"Gene symbol" == "IL18BP")]

```

```{r}
dim(exprs(gset_female))
```


```{r}
gset_female@featureData@data
```


```{r}
gset_female@annotation
```




From GEO2R R Script

```{r}
# Version info: R 3.2.3, Biobase 2.30.0, GEOquery 2.40.0, limma 3.26.8
################################################################
#   Differential expression analysis with limma
library(GEOquery)
library(limma)
library(umap)

# load series and platform data from GEO
gset <- getGEO("GSE48024", GSEMatrix =TRUE, AnnotGPL=TRUE)[[1]]

# make proper column names to match toptable 
fvarLabels(gset_female) <- make.names(fvarLabels(gset_female))

# group membership for all samples
gsms <- "undefined"
sml <- strsplit(gsms, split="")[[1]]

# log2 transformation
ex <- exprs(gset_female)
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) ||
          (qx[6]-qx[1] > 50 && qx[2] > 0)
if (LogC) { ex[which(ex <= 0)] <- NaN
  exprs(gset_female) <- log2(ex) }

# assign samples to groups and set up design matrix
gs <- factor(sml)
groups <- make.names(c("undefined"))
levels(gs) <- groups
gset$group <- gs
design <- model.matrix(~group + 0, gset)
colnames(design) <- levels(gs)

fit <- lmFit(gset, design)  # fit linear model

# set up contrasts of interest and recalculate model coefficients
cts <- paste(groups, c(tail(groups, -1), head(groups, 1)), sep="-")
cont.matrix <- makeContrasts(contrasts=cts, levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)

# compute statistics and table of top significant genes
fit2 <- eBayes(fit2, 0.01)
tT <- topTable(fit2, adjust="fdr", sort.by="B", number=250)

tT <- subset(tT, select=c("ID","adj.P.Val","P.Value","F"))
write.table(tT, file=stdout(), row.names=F, sep="\t")

# Visualize and quality control test results.
# Build histogram of P-values for all genes. Normal test
# assumption is that most genes are not differentially expressed.
tT2 <- topTable(fit2, adjust="fdr", sort.by="B", number=Inf)
hist(tT2$adj.P.Val, col = "grey", border = "white", xlab = "P-adj",
  ylab = "Number of genes", main = "P-adj value distribution")

# summarize test results as "up", "down" or "not expressed"
dT <- decideTests(fit2, adjust.method="fdr", p.value=0.05)

# create Q-Q plot for t-statistic
t.good <- which(!is.na(fit2$F)) # filter out bad probes
qqt(fit2$t[t.good], fit2$df.total[t.good], main="Moderated t statistic")

# volcano plot (log P-value vs log fold change)
colnames(fit2) # list contrast names
ct <- 1        # choose contrast of interest
volcanoplot(fit2, coef=ct, main=colnames(fit2)[ct], pch=20,
  highlight=length(which(dT[,ct]!=0)), names=rep('+', nrow(fit2)))

# MD plot (log fold change vs mean log expression)
# highlight statistically significant (p-adj < 0.05) probes
plotMD(fit2, column=ct, status=dT[,ct], legend=F, pch=20, cex=1)
abline(h=0)

################################################################
# General expression data analysis
ex <- exprs(gset)

# box-and-whisker plot
ord <- order(gs)  # order samples by group
palette(c("#1B9E77", "#7570B3", "#E7298A", "#E6AB02", "#D95F02",
          "#66A61E", "#A6761D", "#B32424", "#B324B3", "#666666"))
par(mar=c(7,4,2,1))
title <- paste ("GSE48024", "/", annotation(gset), sep ="")
boxplot(ex[,ord], boxwex=0.6, notch=T, main=title, outline=FALSE, las=2, col=gs[ord])
legend("topleft", groups, fill=palette(), bty="n")

# expression value distribution
par(mar=c(4,4,2,1))
title <- paste ("GSE48024", "/", annotation(gset), " value distribution", sep ="")
plotDensities(ex, group=gs, main=title, legend ="topright")

# UMAP plot (dimensionality reduction)
ex <- na.omit(ex) # eliminate rows with NAs
ex <- ex[!duplicated(ex), ]  # remove duplicates
ump <- umap(t(ex), n_neighbors = 4, random_state = 123)
par(mar=c(3,3,2,6), xpd=TRUE)
plot(ump$layout, main="UMAP plot, nbrs=4", xlab="", ylab="", col=gs, pch=20, cex=1.5)
legend("topright", inset=c(-0.15,0), legend=levels(gs), pch=20,
col=1:nlevels(gs), title="Group", pt.cex=1.5)
library("maptools")  # point labels without overlaps
pointLabel(ump$layout, labels = rownames(ump$layout), method="SANN", cex=0.6)

# mean-variance trend, helps to see if precision weights are needed
plotSA(fit2, main="Mean variance trend, GSE48024")


```

