---
title: "Load_GSE48024"
author: "Sylvia Yu Zhang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r load}
library(GEOquery)

#library(limma) # Not needed for loading data

#library(umap) # Not needed for loading data

library(knitr) # Needed to render RMD.
```

### Load data.
```{r}
gset <- getGEO("GSE48024", GSEMatrix =TRUE, AnnotGPL=TRUE)
```

Here, "gset" contains two items, one data for male and another data for female. In total, there are 417 samples from 110 female participants, and 431 samples from 166 male participants. 

```{r}
gset_female <- gset[[1]]
gset_female
gset_male <- gset[[2]]
gset_male

# Information given regarding each sample
colnames(gset_female@phenoData)

table(gset_female$`gender:ch1`)
length(unique(gset_female$`subject:ch1`))

table(gset_male$`gender:ch1`)
length(unique(gset_male$`subject:ch1`))
```

Apparently, female and male samples are sequenced using difference version of platforms. Thus, we see different number of probes. 

```{r}
table(gset_female$platform_id) # Illumina HumanHT-12 V4.0 expression beadchip
nrow(gset_female)

table(gset_male$platform_id) # Illumina HumanHT-12 V3.0 expression beadchip
nrow(gset_male)
```

"AnnotGPL" option in "getGEO" function automatically adds annotation corresponding to each gene. However, 16033 probes do not have gene symbols annotated. 

```{r}
# Inforamtion given regarding each probe
colnames(gset_female@featureData)

kable(head(data.frame(ID = gset_female@featureData$"ID",
                      Gene_symbol = gset_female@featureData$"Gene symbol")))

kable(head(data.frame(ID = gset_female@featureData$"ID",
                      GO = gset_female@featureData$"GO:Function")))


sum(gset_female@featureData$"Gene symbol" == "")

```

### Check normalization, and expression range.

Just randomly draw 10 samples for each gender and boxplot each sample. 
```{r}
boxplot(exprs(gset_female)[,sample(1:417, 10, replace=F)], main = "Female")

boxplot(exprs(gset_male)[,sample(1:431, 10, replace=F)], main = "Male")
```

### Number of samples at each time for female and male

```{r}
kable(table(gset_female$`time:ch1`))

kable(table(gset_male$`time:ch1`))
```

### Below shows five probes corresponding to gene "IL18BP", annotation done by "AnnotGPL" option.

I never tried annotation using "AnnotGPL" before, so I'm not sure how it is done exactly. 

```{r}
gset_female@featureData$"ID"[which(gset_female@featureData$"Gene symbol" == "IL18BP")]

```





