---
title: "DE_analysis_male"
author: "Yu (Sylvia) Zhang"
date: "`r Sys.Date()`"
output: 
  html_document:
        toc: true
        toc_float: true
---

```{r load package, message = FALSE}
library(GEOquery)
library(SummarizedExperiment)
library(DESeq2)
library(here)
library(dplyr)
library(limma)
library(nlme)
library(MASS)
library(multcomp)
library(pheatmap)
library(variancePartition)
library(lme4)
library("r2glmm")
```

# Load data

I saved the data into a local directory to save time from downloading each time.

```{r load data, eval = FALSE}
gset <- getGEO("GSE48024", GSEMatrix =TRUE, AnnotGPL=TRUE)
save(gset, file = "/Users/yu.zhang/Desktop/BIOS 784/Data/GSE48024/gset.RData")
```

```{r}
load("/Users/yu.zhang/Desktop/BIOS 784/Data/GSE48024/gset.RData")
```

# Working on male data

```{r}
gset_male <- gset[[2]]
```

```{r}
boxplot(exprs(gset_male)[,sample(1:431, 10, replace=F)])
```

## Subset to probes with gene name

Total 431 samples and 29228 probes with valid gene symbol.

```{r}
gset_male <- gset_male[which(gset_male@featureData@data[["Gene symbol"]] != ""), ]

gset_male
```

## Set up summarized experiment object

```{r}
SE_male  <- SummarizedExperiment(assay = list("exprs" = exprs(gset_male)),
                                  colData = as(gset_male@phenoData, "data.frame"),
                                  rowData = as(gset_male@featureData, "data.frame"))

table(SE_male$`time:ch1`)
```

## Subset to participants sampled at all 4 time points

```{r}
paticipants.all <- as.data.frame(table(SE_male$`subject:ch1`)) %>% filter(Freq == 4)

SE_male <- SE_male[, which(SE_male$`subject:ch1` %in% paticipants.all$Var1)]

table(SE_male$`time:ch1`)

```

## Merge duplicate probes, averaging expression from the same gene

Said from limma for averaging expression: "This function should only be applied to normalized log-expression values, and not to raw unlogged expression values. It will generate an error message if applied to RGList or EListRaw objects."

```{r}
rownames(SE_male)[1:10]

expression.dupRM <- avereps(SE_male@assays@data@listData[["exprs"]], ID=SE_male@elementMetadata@listData[["Gene symbol"]])

# match(colnames(expression.dupRM ) , rownames(SE_male@colData))

SE_male_dupRM <- SummarizedExperiment(assay = list("exprs" = expression.dupRM),
                                  colData = SE_male@colData)

SE_male_dupRM
```

## Remove lower 25% variant genes

```{r}
SE_male_dupRM@metadata[["gene_mean"]] <- rowMeans(SE_male_dupRM@assays@data@listData[["exprs"]])

boxplot(SE_male_dupRM@metadata[["gene_mean"]])

SE_male_dupRM@metadata[["gene_variance"]] <- rowVars(SE_male_dupRM@assays@data@listData[["exprs"]])

boxplot(SE_male_dupRM@metadata[["gene_variance"]])

quantile(SE_male_dupRM@metadata[["gene_mean"]]) # 25% = 7.734803
quantile(SE_male_dupRM@metadata[["gene_variance"]]) # 25% =  0.0022386139, 75% = 0.0134488159

SE_male_dupRM_25up <- SE_male_dupRM[which(SE_male_dupRM@metadata[["gene_variance"]] > 0.0022386139), ]
SE_male_dupRM_25up

#SE_male_dupRM_75up <- SE_male_dupRM[which(SE_male_dupRM@metadata[["gene_variance"]] > 0.0134488159), ]
#SE_male_dupRM_75up

#save(SE_male_dupRM_25up, file = "/Users/yu.zhang/Desktop/BIOS 784/Data/GSE48024/SE_male_dupRM_25up.RData")
#save(SE_male_dupRM_75up, file = "/Users/yu.zhang/Desktop/BIOS 784/Data/GSE48024/SE_male_dupRM_75up.RData")
```

After filtering out lower 25%, we are left with 14692 unique genes and 368 samples.

After filtering out lower 50%, we are left with 9795 unique genes and 368 samples.

After filtering out lower 75%, we are left with 4898 unique genes and 368 samples.

# DE analysis wrt time only

## Function for DE analysis using top 25% most variant genes


```{r}
mixed_test <- function(gene,gene_name, covariates, fml, var_rowname){

  df_test <- cbind(gene , covariates)
  colnames(df_test)[1] <- "gene"
  mix_model <- aov(fml, data=df_test)
  df_ret <- data.frame(unlist(summary(mix_model))["Error: Within.Pr(>F)1"])
  rownames(df_ret) <- var_rowname
  colnames(df_ret) <- gene_name
  return(tibble(df_ret))
}

mixed_test2 <- function(gene,gene_name, covariates, fml, var_rowname){
  
  df_test <- cbind(gene , covariates)
  colnames(df_test)[1] <- "gene"
  mix_model <- aov(fml, data=df_test)
  df_ret <- data.frame(unlist(summary(mix_model))["Error: Within.Pr(>F)2"])
  rownames(df_ret) <- var_rowname
  colnames(df_ret) <- gene_name
  return(tibble(df_ret))
}

```

## Mixed effect model, gene ~ time + 1|ID

#### P-value from mixed effect model

```{r, eval = FALSE}
genes <- rownames(SE_male_dupRM_25up)

cov <- data.frame(ID = SE_male_dupRM_25up$`subject:ch1`,
                  gender = SE_male_dupRM_25up$`gender:ch1`,
                  time = as.character(SE_male_dupRM_25up$`time:ch1`))


mix_test_time <- sapply( 1:nrow(SE_male_dupRM_25up), function(i) mixed_test(   gene = as.data.frame(SE_male_dupRM_25up@assays@data@listData[["exprs"]][which(rownames(SE_male_dupRM_25up) ==  genes[i]),]),
  gene_name = genes[i],
  covariates = cov,
  fml = as.formula(gene ~ time + Error(ID)),
  var_rowname = c("time")) )

mix_test_time <- as.data.frame(unlist(mix_test_time))

colnames(mix_test_time) <- c("p_value_mix_compsym")

write.csv(mix_test_time, file = paste0(here(), "/code/2_DE_analysis/mix_pvalue_compsym.csv"))


```

#### BH-adjusted p-value

```{r}
mix_test_time <- read.csv(file = paste0(here(), "/code/2_DE_analysis/mix_pvalue_compsym.csv"))
rownames(mix_test_time) <- mix_test_time$X

boxplot(mix_test_time$p_value_mix_compsym, main ="P-value from mixed model, gene ~ time + (1|ID)")

# BH adjusted p-value
mix_test_time$BH_pvalue <- p.adjust(mix_test_time$p_value_mix_compsym, method =  "BH", n = length(mix_test_time$p_value_mix_compsym))

boxplot(mix_test_time$BH_pvalue, main ="BH adjusted P-value from mixed model, gene ~ time + 1|ID")

DEgene_pvalue_BH <- mix_test_time %>% as.data.frame() %>% filter(BH_pvalue < 0.01)

nrow(DEgene_pvalue_BH)

```
A total of 4540 DE genes wrt time after BH adjustment.

#### Heatmap on top 50 DE genes from BH adjusted p-values

```{r, fig.height=8, fig.width=6 }

DE_genes <- SE_male_dupRM_25up[which(rownames(SE_male_dupRM_25up) %in% rownames(DEgene_pvalue_BH %>% arrange(BH_pvalue))[1:50]), ]

colData(DE_genes)$time <- factor(colData(DE_genes)$`time:ch1`, levels = c("Day0", "Day1", "Day3", "Day14"))

table(colData(DE_genes)$time)

DE_genes_mean <- sapply(c("Day0", "Day1", "Day3", "Day14"), function(day) rowMeans(assay(DE_genes)[, which(colData(DE_genes)$time == day)]))

pheatmap(DE_genes_mean, cluster_rows = TRUE,  cluster_cols = FALSE, scale = "row", 
          main = "Heatmap on top 50 DE genes wrt time by BH adjusted p-values")

```

### Gene ontology enrichment analysis

```{r, message = FALSE}
library(enrichplot)
library(DOSE)
library(gprofiler2)

DE_gene_name <- gost(query = DEgene_pvalue_BH$X, organism = "hsapiens", multi_query = FALSE,
                     correction_method = "fdr", user_threshold = 0.01, source = "GO")

head(DE_gene_name$result)

table(DE_gene_name[["result"]][["source"]])

#gostplot(DE_gene_name, interactive = TRUE)

colnames(DE_gene_name$result)

gostplot(DE_gene_name, capped = TRUE, interactive = FALSE)

DE_mod = DE_gene_name$result[,c("query",  "term_id",
                                "term_name", "p_value", "query_size", 
                                "intersection_size", "term_size", 
                                "effective_domain_size", "intersection_size")]

DE_mod$"GeneRatio" = paste0(DE_mod$intersection_size,  "/", DE_mod$query_size)
DE_mod$BgRatio = paste0(DE_mod$term_size, "/", DE_mod$effective_domain_size)
DE_mod$"pvalue" <- DE_mod$p_value
DE_mod$Count <- DE_mod$intersection_size
DE_mod$Description<- DE_mod$term_name
DE_mod_enrich  <-  new("enrichResult", result = DE_mod)

dotplot(DE_mod_enrich , x = 'GeneRatio', color = "pvalue")

```

```{r test code, echo=FALSE,eval = FALSE}
library(topGO)
library("AnnotationHub")
library(GO.db)
library(clusterProfiler)
data(geneList)
geneList[1:5]
de <- names(geneList)[1:100]
de[1:5]
x <- enrichDO(de)
dotplot(x)

# Maybe useful? 
ah <- AnnotationHub()
orgs <- subset(ah, ah$rdataclass == "OrgDb")
orgdb <- query(orgs, "Homo sapiens")[[1]]
geneList <- DEgene_pvalue_BH$BH_pvalue
names(geneList) <- DEgene_pvalue_BH$X
GOdata <- new("topGOdata",
    ontology = "BP",
    allGenes = geneList,
    geneSelectionFun = function(x)x,
    annot = annFUN.org, mapping = )

```

# Cluster participants based on diff(day1 - day0) expression 

```{r }

library(mclust)
library(factoextra)

table(SE_male_dupRM_25up$`time:ch1`)

DE10_response <- c("STAT1", "IRF9", "SPI1", "CD74", "HLA-E", "TNFSF13B", "PRDX2", "PRDX3", "E2F2", "PTEN", "ITGB1")

gene_exprs_day1_0 <- list( day1 = SE_male_dupRM_25up[,which(SE_male_dupRM_25up$`time:ch1` == "Day1")],
                           day0 = SE_male_dupRM_25up[,which(SE_male_dupRM_25up$`time:ch1` == "Day0")])

dim(gene_exprs_day1_0[[1]])

gene_exprs_day1_0[[2]] <- gene_exprs_day1_0[[2]][match(rownames(gene_exprs_day1_0[[2]]), rownames(gene_exprs_day1_0[[1]])), match(gene_exprs_day1_0[[2]]$`subject:ch1`, gene_exprs_day1_0[[1]]$`subject:ch1`)]

diff_day1_0 <- gene_exprs_day1_0[["day1"]]@assays@data@listData[["exprs"]] - gene_exprs_day1_0[["day0"]]@assays@data@listData[["exprs"]]

diff_day1_0 <- diff_day1_0[which(rownames(diff_day1_0) %in% DE10_response) , ]

#hist(diff_day1_0[1,], breaks=30)

colnames(diff_day1_0) <- gene_exprs_day1_0[[2]]$`subject:ch1`

GMM <-  Mclust(t(diff_day1_0),  G = 2)
table(GMM$classification)

km<- kmeans(t(diff_day1_0), 2, nstart = 25)
table(km$cluster)

table(GMM$classification , km$cluster)

SE_male_dupRM_25up$response <- GMM$classification[match(SE_male_dupRM_25up$`subject:ch1`, names(GMM$classification)) ]
```

Out of 92 participants, GMM and K-means agree on the classification of 24+49=73 participants.  

For downstream analysis of DE genes on response over time, I'm using result from GMM for now. If we are interested in K-means after I put more visualization, I can definitely switch. 

# DE genes on response over time, using Day 1 - 14 data only

```{r, eval = FALSE}

genes <- rownames(SE_male_dupRM_25up)

expres_test <- SE_male_dupRM_25up[ , which(SE_male_dupRM_25up$`time:ch1` != "Day0")]

expres_test

cov <- data.frame(ID = expres_test $`subject:ch1`,
                  gender = expres_test $`gender:ch1`,
                  time = as.character(expres_test $`time:ch1`),
                  response = factor(expres_test $response))

mix_test_response_time <- sapply( 1:nrow(SE_male_dupRM_25up), function(i) mixed_test2( gene = as.data.frame(expres_test@assays@data@listData[["exprs"]][which(rownames(expres_test) ==  genes[i]),]),
  gene_name = genes[i],
  covariates = cov,
  fml = as.formula(gene ~ time*response + Error(ID)),
  var_rowname = c("response_time")) )

mix_test_response_time <- as.data.frame(unlist(mix_test_response_time))

colnames(mix_test_response_time) <- c("p_value_mix_compsym")

write.csv(mix_test_response_time, file = paste0(here(), "/code/2_DE_analysis/mix_response_time_pvalue_compsym.csv"))

```


#### BH-adjusted p-value

57 DE genes on response over time for BH p-value cutoff = 0.05.

340 DE genes on response over time for BH p-value cutoff = 0.1.

```{r}
mix_test_response_time <- read.csv(file = paste0(here(), "/code/2_DE_analysis/mix_response_time_pvalue_compsym.csv"))
rownames(mix_test_response_time) <- mix_test_response_time$X

boxplot(mix_test_response_time$p_value_mix_compsym, main ="P-value from mixed model, gene ~ time*response + (1|ID)")

# BH adjusted p-value
mix_test_response_time$BH_pvalue <- p.adjust(mix_test_response_time$p_value_mix_compsym, method =  "BH", n = length(mix_test_response_time$p_value_mix_compsym))

boxplot(mix_test_response_time$BH_pvalue, main ="BH adjusted P-value from mixed model, gene ~ time*response + 1|ID")

DEgene_pvalue_BH_response <- mix_test_response_time %>% as.data.frame() %>% filter(BH_pvalue < 0.1)

nrow(DEgene_pvalue_BH_response)

```


#### Heatmap, this is interesting


```{r, fig.height=8, fig.width=8}

DE_genes_response <- SE_male_dupRM_25up[which(rownames(SE_male_dupRM_25up) %in% rownames(DEgene_pvalue_BH_response%>% arrange(BH_pvalue))[1:50]), which(SE_male_dupRM_25up$`time:ch1` != "Day0")]

colData(DE_genes_response)$time <- factor(colData(DE_genes_response)$`time:ch1`, levels = c("Day0", "Day1", "Day3", "Day14"))

table(colData(DE_genes_response)$time)

DE_genes_response_mean <- sapply(c( "Day1", "Day3", "Day14"), function(day) rowMeans(assay(
DE_genes_response)[, which(colData(
DE_genes_response)$time == day)]))

pheatmap(DE_genes_response_mean, cluster_rows = TRUE,  cluster_cols = FALSE, scale = "row", 
          main = "Heatmap on top 50 DE genes wrt response over time by BH adjusted p-values")

```

### Gene ontology enrichment analysis

```{r, message = FALSE}

DE_gene_response_name <- gost(query = DEgene_pvalue_BH_response$X, organism = "hsapiens", multi_query = FALSE,
                     correction_method = "fdr", user_threshold = 0.01, source = "GO")

head(DE_gene_response_name$result)

table(DE_gene_response_name[["result"]][["source"]])

colnames(DE_gene_response_name$result)

gostplot(DE_gene_response_name, capped = TRUE, interactive = FALSE)

DE_mod_response = DE_gene_response_name$result[,c("query",  "term_id",
                                "term_name", "p_value", "query_size", 
                                "intersection_size", "term_size", 
                                "effective_domain_size", "intersection_size")]

DE_mod_response$"GeneRatio" = paste0(DE_mod_response$intersection_size,  "/", DE_mod_response$query_size)
DE_mod_response$BgRatio = paste0(DE_mod_response$term_size, "/", DE_mod_response$effective_domain_size)
DE_mod_response$"pvalue" <- DE_mod_response$p_value
DE_mod_response$Count <- DE_mod_response$intersection_size
DE_mod_response$Description<- DE_mod_response$term_name
DE_mod_response_enrich  <-  new("enrichResult", result = DE_mod_response)

dotplot(DE_mod_response_enrich , x = 'GeneRatio', color = "pvalue")

```